Realizing Temporal Graphs

### 2023-01-09 RealizingTemporalGraphs.tex

\todo[inline]{NK:Following definition and observation are copied from the previous section (need to delete from one part).}

@@

We construct a labelling $\lambda$ as follows.
\begin{itemize}
    \item All edges incident with a clause vertex $c$ obtain label one.
    %\item For each variable $x$, we set $\lambda(\{x^T,x^F\})=1$.
    \item If variable $x$ is set to \texttt{true}, we set $\lambda(\{x^F,v\})=3$.
    \item If variable $x$ is set to \texttt{false}, we set $\lambda(\{x^T,v\})=3$.
    \item We set the labels of all other edges to two.
\end{itemize}

@@

Observe that edges incident with $y^T, y^F$ that have label three connect $y^T, y^F$ to $v$. Furthermore, $v$ has distance two from $x$ in the unlabelled graph $G$. Since all edges incident with variable vertices $x$ have label two, there is no temporal path starting at $x$ and arriving at $v$ at time two, that could then continue to $y_T$ or $y_F$ using an edge with label three. Hence, there is no temporal path from $x$ to $y^T, y^F$ using an edge with label three. Since all edges incident with $x$ have label two and variable vertices $x$ are non-adjacent to variable vertices $y^T,y^F$, there cannot be a temporal path from variable vertex $x$ to variable vertex $y^T$ or $y^F$.
\item For all variable vertices $x$ and clause vertices $c$ we have $d(v,x)=d(v,c)=\infty$:
\todo[inline]{New matrix entries D for period Delta: d(v,x)=Delta (using two edges, with labels L+1,L). 
New matrix entries D for period Delta: d(v,c)=Delta (using two edges, with labels L,L-1).}

@@

We now introduce an algorithm that determines the labeling of $G$, satisfying $D$, that runs in $f(k) \cdot \text{poly}(n)$ time.
Our algorithm performs the following steps.
\begin{enumerate}
    \item Create $G'$ and $D'$ from $G$ and $D$ respectively, by removing all degree $1$ vertices from $G$ and $D$.
    \item Compute a minimum feedback edge set $F$ in $G'$ (note $F$ is also a minimum feedback edge set of $G$).
    \item Determine vertices of interest $U$ of the graph $G'$.
    \item Find the (partial) labelling $\lambda$ of $G'$, that connects all pairs of vertices from $U$, satisfying their respective values in $D'$.
    \item Extend the partial labelling $\lambda$ to the labeling of the whole $G'$, respecting $D'$.
    \item Extend the labelling $\lambda$ of $G'$ to the labeling of $G$, respecting $D$.
\end{enumerate}
For each of the steps we now argue its correctness and time complexity.

\paragraph{Step 1.}
In the first step of the algorithm we create an instance $D'$, with the underlying graph $G'$, from $D$ and $G$, by simply removing all the vertices of degree one from $G$ and their corresponding rows and columns from the matrix $D$. 
This step is executed in $O(n+m)$ time.
Since we determine the labels of edges in $G \setminus G'$ in the last step, the the whole solution is not affected by performing this step.

\paragraph{Step 2.}
To find a minimum feedback edge set $F$ of $G'$ we find a spanning tree $T$ of $G'$ and get $F = E(G) \setminus E(T)$. 
Clearly we perform this step in linear time.

\paragraph{Step 3.}
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subset V(G')$ be the set of leaves in $T$, $V_2 \subset V(G')$ be the set of vertices of degree two in $T$, that are incident to some edges in $F$, and let $V_3 \subset V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ is incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with $U = V_1 \cup V_2 \cup V_3$ the set of \emph{vertices of interest}. It follows that $|U| \leq 4k$.
Clearly we perform this step in linear time.

\paragraph{Step 4}
In this step we want to determine the labeling of fastest temporal paths among all vertices in $U$.
For a pair $u,v$ of vertices in $U$, there are $2^{O(k)}$ possible paths in $G'$ between them, where each one could be the fastest temporal path from $u$ to $v$ and $v$ to $u$. Therefore we get $2^{O(k)}$ partial (different) labelings of $G'$. \todo{NK: How??}
This is true for any pair of vertices in $U$. 
We then try to join the labelings of different pairs of vertices and end up with $(2^{O(k)})^{k^2} = 2^{O(k^3)}$ different labelings.
Some of them we can reject immediately, namely if two different labellings label the same edge with the different label.
At the end we end up with $22^{O(k^3)}$ partial labelings $\lambda$ of $G'$, where $(G',\lambda)$ has labels on all fastest paths among any pair of vertices in $U$,  and the duration of the fastest temporal path from $u$, to $v$, where $u,v \in U$ satisfies the value $D_{u,v}$. 
\todo[inline]{Missing details.}

\paragraph{Step 5.}
In the partially labeled temporal graph $(G',\lambda)$ the durations of the fastest temporal paths among vertices in $U$ satisfy the conditions of $D'$,
but there are still some labels without labels.
Let $e=(uv)$ be one such edge.
\todo[inline]{TBD}

\paragraph{Step 6.}
Let $\lambda$ be the labelling of $G'$, that satisfies $D'$.
Let $s \in V(G) \setminus V(G')$, be a vertex of degree one, that was removed from $G$ in Step 1,
and let $u$ be its neighbor in $G$.
First, we can easily assume that $|N_{G'}(u)|\geq 1$, if this is not true, then either $G$ is a star or $G'$ is disconnected and therefore also $G$ is disconnected, where $u$ is in a part of $G$ that forms a star.
We can solve the problem for stars in polynomial time (see \cref{thm:deltaExact-PolyTimeTrees}).

@@

    \begin{claim}
    Labelling $\lambda$ is a solution for the \textsc{Exact Realization $(\Delta)$} if and only if
    for every cycle $C$ of $(G,\lambda)$ it holds that the sum of the travel delays along $C$ is $0$ modulo $\Delta$.
    \end{claim}

>>>>

@@

We construct a labeling $\lambda$ as follows.
\begin{itemize}
    \item All edges incident with a clause vertex $c$ obtain label one.
    %\item For each variable $x$, we set $\lambda(\{x^T,x^F\})=1$.
    \item If variable $x$ is set to \texttt{true}, we set $\lambda(\{x^F,v\})=3$.
    \item If variable $x$ is set to \texttt{false}, we set $\lambda(\{x^T,v\})=3$.
    \item We set the labels of all other edges to two.
\end{itemize}

@@

Observe that edges incident with $y^T, y^F$ that have label three connect $y^T, y^F$ to $v$. Furthermore, $v$ has distance two from $x$ in the unlabeled graph $G$. Since all edges incident with variable vertices $x$ have label two, there is no temporal path starting at $x$ and arriving at $v$ at time two, that could then continue to $y_T$ or $y_F$ using an edge with label three. Hence, there is no temporal path from $x$ to $y^T, y^F$ using an edge with label three. Since all edges incident with $x$ have label two and variable vertices $x$ are non-adjacent to variable vertices $y^T,y^F$, there cannot be a temporal path from variable vertex $x$ to variable vertex $y^T$ or $y^F$.
\item For all variable vertices $x$ and clause vertices $c$ we have $d(v,x)=d(v,c)=\infty$:
\todo[inline]{New matrix entries D for period Delta: d(v,x)=Delta (using two edges, with labels L+1,L). 
New matrix entries D for period Delta: d(v,c)=Delta (using two edges, with labels L,L-1).}

@@

We now introduce an algorithm that determines the labeling of $G$, satisfying $D$, that runs in $f(k) \cdot \text{poly}(n)$ time.
Our algorithm performs the following steps.
\begin{enumerate}
    \item Create $G'$ and $D'$ from $G$ and $D$ respectively, by removing all degree $1$ vertices from $G$ and $D$.
    \item Compute a minimum feedback edge set $F$ in $G'$ (note, $F$ is also a minimum feedback edge set of $G$).
    \item Determine vertices of interest $U$ of the graph $G'$.
    \item \label{enum:FPT-labelingVerticesOfInterest} Guess the (partial) labeling $\lambda$ of $G'$, that connects all pairs of vertices from $U$, satisfying their respective values in $D'$.
    \item \label{enum:FPT-labelingraphNoLeafs} Extend the partial labeling $\lambda$ to the labeling of the whole $G'$, respecting $D'$.
    \item \label{enum:FPT-labelingWholeGraph} Extend the labeling $\lambda$ of $G'$ to the labeling of $G$, respecting $D$.
\end{enumerate}
It is important to point out, that in step~\ref{enum:FPT-labelingVerticesOfInterest}, our algorithm does not build a labeling satisfying conditions in this step,
and then tries to extend it in steps~\ref{enum:FPT-labelingraphNoLeafs} and \ref{enum:FPT-labelingWholeGraph},
but it
builds an integer linear program (ILP) problem with $O(k)$ variables,
with restrictions satisfying condition from steps~\ref{enum:FPT-labelingVerticesOfInterest}--\ref{enum:FPT-labelingWholeGraph}.
For each of the steps we now argue its correctness and time complexity.

\subsection{Steps 1 -- 3 of the FPT algorithm}
In the first step of the algorithm we create an instance $D'$, with the underlying graph $G'$, from $D$ and $G$, by simply removing all the vertices of degree one from $G$ and their corresponding rows and columns from the matrix $D$. 
This step is executed in $O(n+m)$ time.
Since we determine the labels of edges in $G \setminus G'$ in the last step, the whole solution is not affected by performing this step.

In the second step,
we determine a minimum feedback edge set $F$ of $G'$ by finding a spanning tree $T$ of $G'$ and set $F = E(G) \setminus E(T)$. 
We can perform this step in linear time.

In the third step we determine the set $U$ of \emph{vertices of interest}.
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subset V(G')$ be the set of leaves in $T$, $V_2 \subset V(G')$ be the set of vertices of degree two in $T$, that are incident to some edges in $F$, and let $V_3 \subset V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ are incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with $U = V_1 \cup V_2 \cup V_3$ the set of vertices of interest. It follows that $|U| \leq 4k$.
We can determine vertices of $U$ in linear time.

\subsection{Step 4 of the FPT algorithm}
In this step we want to guess the structure of the fastest paths among all pairs of vertices of $U$. %and then determine their labeling.
For a pair $u,v$ of vertices in $U$, there are $2^{O(k)}$\todo{NK: why? \\ Isn't $k!$} possible paths in $G'$ between them, where each one could be the fastest temporal path from $u$ to $v$ and $v$ to $u$.

We claim that knowing the structure of the fastest temporal path among two vertices, determines the labeling of such path, up to the label of the first edge.
To prove this we use the following observation, that follows from the definition of the duration of the fastest temporal path.
\begin{observation}\label{obs:FirstLabelAndDuration}
    Let $P$ be the fastest temporal path from $u$ to $v$, where $e_1, e_p \in P$ are its first and last edge, respectively.
    Then knowing the label of the first (last) edge and the duration of $P$, one can uniquely determine the label of the last (first) edge of $P$.
\end{observation}
This is true as, by the definition the duration of $P$ is calculated as the difference of labels of last and first edge plus $1$,
where the label of last edge is considered with some delta periods,
\ie $d(P) = p_i \Delta + \lambda(e_p) - \lambda (e_1) + 1$.
Therefore $d(P) \pmod \Delta \equiv  (\lambda(e_p) - \lambda (e_1) + 1) \pmod \Delta$.
We can now state and prove the following claim.
\begin{claim}
    Suppose $P = (u=v_1, v_2, \dots, v_p = v)$, where $p \geq 2$, is a $u,v$ path in $G'$, which is also the underlying path of the fastest temporal path from $u$ to $v$.
    One can determine the labelling $\lambda$ of $P$, up to the constant $c_{uv} = \lambda(uv_2)$,
    such that the labelling $\lambda$ respects $D'$.
\end{claim}
\begin{proof}
Let $P$ be the fastest temporal path from $u$ to $v$.
We have the following two options:
\begin{enumerate}
\item no other vertex from $P$ is a vertex of interest, \ie $P \setminus \{u,v\} \cap U = \emptyset $,
\item there are at least $3$ vertices of interest in $P$.
\end{enumerate}

In the first case we get that all other vertices in $P$ are reached from $u$ using a (sub)path of $P$.
To prove this suppose for the contradiction, that there is a vertex $v_i \in P \setminus \{u,v\}$, that is reached from $v$ on a path different than $P_i = (u, v_2, v_3, \dots, v_i)$.
Since only $u$ and $v$ are the vertices of interest, it follows that all other vertices on $P$ are of degree $2$. 
Then, the only way to reach $v_i$ from $v$, that differs from $P$, would be to go from $v$ to $u$ using a differnt path $P_2$,
and then go from $u$ to $v_{p-1}, v_{p-2}, \dots, v_i$.
But since $P$ is the fastest temporal path from $u$ to $v$, we get that $d(P_2) \geq d(P)$ and $d(P_2 + (u,v_i) > d(P) > d(P_i)$.

Now to label $P$ we use the fact that the fastest temporal path from $u$ to any $v_i \in P$ is a subpath of $P$, 
therefore we can label each edge using the \cref{obs:FirstLabelAndDuration},
where the duration from $u$ to $v_i$ equals to $D_{u,v_i}$ and 
we set the first label of $P$ to be a constant $c_{uv}$.
This gives us a condition on the label of each edge of $P$, that depends on the value $c_{u,v}$.

Let us now prove the correctness of the second case.
We denote with $v_i \in U$ a vertex of interest in $P \setminus \{u,v\}$.
There are two options, either $v_i$ is reached from $u$ using the subpath $P_i = (u, v_2, v_3, \dots, v_i)$ of $P$,
or there exists a fastest temporal path $P' = (u, w_2, w_3, \dots, w_{p'} = v_i)$, for which $P \cap P' = \{u, v_i\}$.
If the temporal path from $u$ to $v_i$ is a subpath of $P$, then we determine the labeling the same as in previous case.
So we can suppose that the fastest temporal path from $u$ to $v_i$ is of form $P'$.\todo{TBD}
\end{proof}

Therefore, knowing the structure of the exact temporal path $P$ from $u$ to $v$, labeling the path equals to creating a variable $c_{u,v} \in [\Delta]$ that represents the label of the first edge of $P$, 
and $O(len(P))$ constraints, that uniquely determine the labels of all other edges of $P$.
This holds for all pairs of vertices in $U$, therefore
knowing the structure of fastest temporal paths among all pairs of vertices in $U$ creates $k^2$ variables of form $c_{u,v}$.

There are $(2^{O(k)})^{k^2} = 2^{O(k^3)}$ different possible fastest temporal paths among all pairs of vertices in $U$. 
Therefore we have to build $2^{O(k^3)}$ different ILP problems.

\subsection{Step 5 of the FPT algorithm}
In the partially labeled temporal graph $(G',\lambda)$ the durations of the fastest temporal paths among vertices in $U$ satisfy the conditions of $D'$,
but there are still some labels without labels.
Let $e=(uv)$ be one such edge.
\todo[inline]{TBD}

\subsection{Step 6 of the FPT algorithm}
\todo[inline]{This section is wrong, we keep this here as we will use some of the ideas.}
Let $\lambda$ be the labeling of $G'$, that satisfies $D'$.
Let $s \in V(G) \setminus V(G')$, be a vertex of degree one, that was removed from $G$ in Step 1,
and let $u$ be its neighbor in $G$.
First, we can easily assume that $|N_{G'}(u)|\geq 1$, if this is not true, then either $G$ is a star or $G'$ is disconnected and therefore also $G$ is disconnected, where $u$ is in a part of $G$ that forms a star.
We can solve the problem for stars in polynomial time (see \cref{thm:deltaExact-PolyTimeTrees}).

@@

    \begin{claim}
    labeling $\lambda$ is a solution for the \textsc{Exact Realization $(\Delta)$} if and only if
    for every cycle $C$ of $(G,\lambda)$ it holds that the sum of the travel delays along $C$ is $0$ modulo $\Delta$.
    \end{claim}
